---
unit_title: purrr inside mutate
needs: [purrr-basics, function-vector]
theme: program
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(tibble.print_min = 5, tibble.print_max = 5)
library(tidyverse)
```

In the vector functions unit, you learned to add new variables to tibbles by giving `mutate()` a vector with an element for each row in the tibble. You saw that you can do any of the following to create this vector:

* Give `mutate()` a single value, which is then repeated for each row in the tibble.
* Explicitly give `mutate()` a vector with an element for each row in the tibble.
* Create a vector using a vector function like `+` or `case_when()`.

In this reading, you'll learn about a fourth option: supplying `mutate()` with a vector created by a `map` function.

## Using `map` functions inside `mutate()`

Here's a simple scalar function that turns feelings into emoticons:

```{r}
emoticons_1 <- function(x) {
  if (x == "happy") {
    ":)"
  } else if (x == "sad") {
    ":("
  } else {
    "Feeling too complex for emoticons"
  }
}
```

Recall that scalar functions fail if you try to give them vectors of values:

```{r}
feelings <- c("happy", "befuddled")
emoticons_1(feelings)
```

As you already know, we can turn `emoticons()` into a vector function by rewriting it using `if_else()`, `case_when()`, or `recode()`. Another option, however, is to leave `emoticons()` as a scalar function and apply it to the elements of `feelings` one-by-one using a `map` function:

```{r}
map_chr(feelings, emoticons_1)
```

Here, we place each feeling in `feelings` on the conveyor belt. The worker picks each feeling up, finds the appropriate emoticon, and then places the emoticon on the output belt. 

If we want to use `emoticons()` to add a variable to a tibble, we can place this `map` call inside `mutate()`:

```{r}
df <- 
  tibble(
    feeling = c("sad", "compunctious", "happy")
  ) 

df %>% 
  mutate(emoticon = map_chr(feeling, emoticons_1))
```

Remember that when you reference an existing column inside `mutate()`, you reference the entire vector of values. Therefore, the call to `map_chr()` works inside `mutate()` just as it does outside, applying `emoticons()` to each element of the `feeling` column and producing a vector with an emoticon for each feeling in the tibble.

This pattern will work with any scalar function, but remember to think about which `map` suffix to use. If the output type of the scalar function doesn't match the `map` suffix, you'll get an error about coercion: 

```{r, error=TRUE}
df %>% 
  mutate(emoticon = map_int(feeling, emoticons_1))
```

## Additional arguments and anonymous values

You can use the `map` functions inside `mutate()` as you would outside `mutate()`, and so you can still specify additional arguments by listing them after the function:

```{r}
emoticons_2 <- function(x, default = "???") {
  if (x == "happy") {
    ":)"
  } else if (x == "sad") {
    ":("
  } else {
    default
  }
}

df %>% 
  mutate(emoticon = map_chr(feeling, emoticons_2, default = NA_character_))
```

as well as create anonymous functions:

```{r}
df %>% 
  mutate(is_happy = map_chr(feeling, ~ str_c("I feel ", emoticons_2(.))))
```

