---
unit_title: "Spatial visualization"
theme: visualize
needs: [spatial-basics]
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Setup

We'll start by loading the tidyverse, sf, and read in a couple of sample datasets.

```{r setup, message = FALSE}
library(tidyverse)
library(sf)

nc <- read_sf(system.file("shape/nc.shp", package = "sf"))
states <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))
```

## `geom_sf()`

The easiest way to get started is to supply an sf object to `geom_sf()`:

```{r}
ggplot() +
  geom_sf(data = nc)
```

Notice that ggplot2 takes care of setting the aspect ratio correctly.

You can supply other aesthetics: for polygons, `fill` is most useful:

```{r}
ggplot() +
  geom_sf(aes(fill = AREA), data = nc, color = "white")
```

When you include multiple layers, ggplot2 will take care of ensuring that they all have a common coordinate reference system (CRS) so that it makes sense to overlay them.

```{r}
ggplot() +
  geom_sf(data = states) + 
  geom_sf(data = nc)
```

You can combine `geom_sf()` with other geoms. In this case, `x` and `y` positions are assumed be in the same CRS as the sf object (typically these will be longitude and latitude). 

```{r}
ggplot() +
  geom_sf(data = nc) +
  annotate(geom = "point", x = -80, y = 35, color = "red", size = 4)
```

## `coord_sf()`

You'll need to use `coord_sf()` for two reasons:

* You want to zoom into a specified region of the plot by using `xlim` and
  `ylim`
    
    ```{r}
    ggplot() +
      geom_sf(data = nc) +
      annotate(geom = "point", x = -80, y = 35, color = "red", size = 4) + 
      coord_sf(xlim = c(-81, -79), ylim = c(34, 36))
    ```
  
* You want to override to use a specific projection. If you don't specify the
  `crs` argument, it just uses the one provided in the first layer. The following example uses an Albers Equal Area projection and the NAD83 datum, with an EPSG code of 102003.
  
    ```{r}
    ggplot() +
      geom_sf(data = states) +
      coord_sf(crs = st_crs(102003))
    ```
