---
unit_title: "Vector and summary functions"
theme: wrangle
needs: [manip-basics, missing-values]
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 3)
set.seed(1014)
```

```{r setup, message=FALSE}
library(tidyverse)
library(nycflights13)
```

`mutate()` and `summarise()` operate on data frames. You use them with vector and summary functions which work with individual variables (or vectors).

## Vector functions

A vector function takes one (or sometimes more) vectors as input and returns a vector of the same length as output. These are typically used with `mutate()` to create new variables.

I can't list every vector function, but here's a few that are often useful:

*   Arithmetic operators: `+`, `-`, `*`, `/`, `^`. These are all vectorised,
    using the so called "recycling rules". If one parameter is shorter than 
    the other, it will be automatically extended to be the same length. This 
    is most useful when one of the arguments is a single number: `air_time / 60`,
    `hours * 60 + minute`, etc.
    
    Arithmetic operators are also useful in conjunction with the aggregate
    functions you'll learn about later. For example, `x / sum(x)` calculates 
    the proportion of a total, and `y - mean(y)` computes the difference from 
    the mean.
    
*   Modular arithmetic: `%/%` (integer division) and `%%` (remainder), where
    `x == y * (x %/% y) + (x %% y)`. Modular arithmetic is a handy tool because 
    it allows you to break integers up into pieces. For example, in the 
    flights dataset, you can compute `hour` and `minute` from `dep_time` with:
    
    ```{r}
    flights %>% 
      transmute(
        dep_time,
        hour = dep_time %/% 100,
        minute = dep_time %% 100
      )
    ```
  
*   Logs: `log()`, `log2()`, `log10()`. Logarithms are an incredibly useful
    transformation for dealing with data that ranges across multiple orders of
    magnitude. They also convert multiplicative relationships to additive, a
    feature we'll come back to in modelling.
    
    All else being equal, I recommend using `log2()` because it's easy to
    interpret: a difference of 1 on the log scale corresponds to doubling on
    the original scale and a difference of -1 corresponds to halving.

*   Logical comparisons, `<`, `<=`, `>`, `>=`, `!=`, which you learned about
    earlier. If you're doing a complex sequence of logical operations it's 
    often a good idea to store the interim values in new variables so you can
    check that each step is working as expected.

*   `if_else()` allows you perform a conditional calculation. The first argument
    should be a logical statement; the second argument is the value to use if
    the first argument is true; the third argument is the value to use if the 
    first argument is false.

    ```{r}
    flights %>% 
      transmute(
        on_time = if_else(arr_delay < 0, "early", "late")
      )
    ```

*  `case_when()` allows you to combine multiple logical conditions, evaluated
   in turn. Each condition goes on the left hand side of the `~`, and the 
   result goes on the right hand side.
   
    ```{r}
    flights %>% 
      transmute(
        on_time = case_when(
          abs(arr_delay) < 10 ~ "on time",
          arr_delay < 0 ~ "early",
          arr_delay > 0 ~ "late",
          is.na(arr_delay) ~ "cancelled"
        )
      )
    ```
   

## Summary functions

A summary function takes a vector of inputs and returns a single output. They are most commonly used with `summarise()`.

Just using means, counts, and sum can get you a long way, but R provides many other useful summary functions:

*   Measures of location: we've used `mean(x)`, but `median(x)` is also
    useful. The mean is the sum divided by the length; the median is a value 
    where 50% of `x` is above it, and 50% is below it.
    
    It's sometimes useful to combine aggregation with logical subsetting. 
    We haven't talked about this sort of subsetting yet, but you'll learn more
    about it later.
    
    ```{r}
    flights %>% 
      group_by(year, month, day) %>% 
      summarise(
        avg_delay1 = mean(arr_delay, na.rm = TRUE),
        # the average positive delay
        avg_delay2 = mean(arr_delay[arr_delay > 0], na.rm = TRUE)
      ) %>% 
      ungroup()
    ```

*   Measures of spread: `sd(x)`, `IQR(x)`, `mad(x)`. The mean squared deviation,
    or standard deviation or sd for short, is the standard measure of spread.
    The interquartile range `IQR()` and median absolute deviation `mad(x)`
    are robust equivalents that may be more useful if you have outliers.
    
    ```{r}
    # Why is distance to some destinations more variable than to others?
    flights %>% 
      group_by(dest) %>% 
      summarise(distance_sd = sd(distance, na.rm = TRUE)) %>% 
      arrange(desc(distance_sd))
    ```
  
*   Measures of rank: `min(x)`, `quantile(x, 0.25)`, `max(x)`. Quantiles
    are a generalisation of the median. For example, `quantile(x, 0.25)`
    will find a value of `x` that is greater than 25% of the values,
    and less than the remaining 75%.

    ```{r}
    # When do the first and last flights leave each day?
    flights %>% 
      group_by(year, month, day) %>% 
      summarise(
        first = min(sched_dep_time, na.rm = TRUE),
        last = max(sched_dep_time, na.rm = TRUE)
      ) %>% 
      ungroup()
    ```
  
*   Measures of position: `first(x)`, `nth(x, 2)`, `last(x)`. These work 
    similarly to `x[1]`, `x[2]`, and `x[length(x)]` but let you set a default 
    value if that position does not exist (i.e. you're trying to get the 3rd
    element from a group that only has two elements). For example, we can
    find the first and last departure for each day:
    
    ```{r}
    flights %>% 
      filter(!is.na(dep_time)) %>% 
      group_by(year, month, day) %>% 
      summarise(
        first_dep = first(dep_time), 
        last_dep = last(dep_time)
      ) %>% 
      ungroup()
    ```
    
    These functions are complementary to filtering on ranks. Filtering gives
    you all variables, with each observation in a separate row:
    
    ```{r}
    flights %>% 
      filter(!is.na(dep_time)) %>% 
      group_by(year, month, day) %>% 
      mutate(r = min_rank(desc(dep_time))) %>% 
      filter(r %in% range(r)) %>% 
      ungroup()
    ```

*   Counts: You've seen `n()`, which takes no arguments, and returns the 
    size of the current group. To count the number of non-missing values, use
    `sum(!is.na(x))`. To count the number of distinct (unique) values, use
    `n_distinct(x)`.
    
    ```{r}
    # Which destinations have the most carriers?
    flights %>% 
      group_by(dest) %>% 
      summarise(carriers = n_distinct(carrier)) %>% 
      arrange(desc(carriers))
    ```
    
*   Counts and proportions of logical values: `sum(x > 10)`, `mean(y == 0)`.
    When used with numeric functions, `TRUE` is converted to 1 and `FALSE` to 0. 
    This makes `sum()` and `mean()` very useful: `sum(x)` gives the number of 
    `TRUE`s in `x`, and `mean(x)` gives the proportion.
    
    ```{r}
    # How many flights left before 5am? (these usually indicate delayed
    # flights from the previous day)
    flights %>% 
      filter(!is.na(dep_time)) %>%
      group_by(year, month, day) %>% 
      summarise(n_early = sum(dep_time < 500)) %>% 
      ungroup()
    
    # What proportion of flights are delayed by more than an hour?
    flights %>% 
      filter(!is.na(arr_delay)) %>%
      group_by(year, month, day) %>% 
      summarise(hour_perc = mean(arr_delay > 60)) %>% 
      ungroup()
    ```
